Characteristic polynomials of p-adic matrices

### 2017-02-04 charpoly/charpoly.tex

\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{Soft-Oh}

@@

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over $\OK / \pi\OK$.
\end{proof}

Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and
let $\alpha = \max(\lvert \det(M)^{-1} \rvert, 1)$.  \todo{What if $\det(M) = 0$?}. Then, for all $\rho \in (0, 1]$ and all
$r \in (0, \alpha \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
\end{prop}

@@

Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
Since the precision polygon lies below the Newton polygon of $M$ \cite{caruso-roe-vaccon:15a}*{Prop. 3.4},
and since the smallest entry of the Newton polygon occurs either with the leading
coefficient or the constant term, $B_W(1) \subset d\chi(M)(B_V(\alpha))$.

@@

\begin{proof}
Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
\pmod{\chi_M}$, we deduce $(X{-}N) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}N$
modulo $\chi_M$. On the other hand, a direct computation shows that
every column vector $W$ lying in the right kernel of $X{-}N$ modulo 
$\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
K[X]/\chi_M$. We deduce that $C \equiv P \cdot V^t B \pmod{\chi_M}$
for some row vector $B$.
Applying the same reasoning with $M^t$, we find that $B$ can be
written $B = \alpha V Q^t$ for some $\alpha \in K[X]/\chi_M$ and
we are done.
\end{proof}

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ coefficients to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Subsection \ref{ssec:eigenvalues} that interesting
informations can be read off on this short form $(\alpha, P, Q, 
\chi_M)$.

@@

In this section, we combine the computation of an Hessenberg form
of a matrix and the computation of the inverse through the SNF
over a complete discrete valuation field
to compute $\com (X-M)$ and $d \chi$
in a way such that if $M \in M_n(O_K),$
only division by invertible elements of $O_K$
occur.
%The computation of an Hessenberg form provides an $O(n^3)$
%algorithm to compute the characteristic polynomial.

@@

We begin with the study of the computation of
an approximate Hessenberg form.

\begin{deftn}
$M \in M_n (k)$ is said to be an Hessenberg matrix if
\[\forall i \in \llbracket 3, n \rrbracket, 
\forall j \in \llbracket 1, i-2 \rrbracket, M_{i,j}=0. \]
If $M \in M_n (k)$ and $H \in M_n (k)$ is an Hessenberg 
matrix such that there exists $P \in GL_n(k)$
with $H=PMP^{-1},$ we say that H is an Hessenberg form
of $M.$
\end{deftn}

\begin{deftn}
In the previous definition of an Hessenberg matrix $M,$
 if we only have $M_{i,j}= O(\pi^{n_{i,j}}),$ then we say
 that $M$ is an approximate Hessenberg matrix.
 This generalize directly to approximate Hessenberg form. 
\end{deftn}

It is not hard to prove that every matrix over a field admits
an Hessenberg form.
We prove here that over $K,$ if a matrix is 
known at finite (jagged) precision,
we can compute an approximate Hessenberg form of it.
Moreover, we can provide an exact change of basis matrix.
It relies on the following algorithm.

@@

\noindent 2.\  \:  {\bf swap} the row $j+1$ with a row $i_{min}$ ($i_{min} \geq 2$) s.t. $val(H_{i_{min},j})$ is minimal.

@@

In this Subsection, we prove that to compute the inverse of
a matrix over a CDVF $K$, SNF is precision-wise optimal in the flat-precision case.
We first recall what is the differential of the computation of the inverse of a matrix.

@@

In this subsection, we want to compute $\com (X Id -H)$
by using the SNF computation of the previous subsection.
As it is, in a way, optimal for the computation of the
inverse in a CDVF, we hope to get reasonnable behaviour
for the computation of $\com (X-M)$ even though the coefficients
are not, \textit{a priori}, set to live in a CDVF.
We show that we can reduce this task to the computation
of an SNF over a CDVF, with no division in $K.$

@@

\begin{rem}
The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
division free algorithm of \cite{}. It is likely that, proceeding so, we 
limit the loss of precision.
\end{rem}

>>>>

\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{$\softO$, $\omega$, scalar matrices and $I_n$}

@@

\begin{prop}
\label{prop:intsurj}
For $M \in M_n(\OK)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surjp} the image of $M_n(\OK)$ under $d\chi_M$ is $\OK[X] \cap K_{<n}[X]$.
\item \label{pt:cycp} the reduction of $M$ modulo $\pi$ has a cyclic vector
\end{enumerate}
\end{prop}
\begin{proof}
The condition \eqref{pt:surjp} is equivalent to the surjectivity of $d\chi_M$ modulo $\pi$.  The equivalence with
\eqref{pt:cycp} follows the same argument as Proposition \ref{prop:surjectivity}, but over the residue field of $K$.
\end{proof}

Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$, and let
$\sigma_1(M), \dots, \sigma_n(M)$ denote the elementary divisors of $M$.

\begin{prop} \label{prop:mainlem}
Suppose that $M \in M_n(K)$ satisfies one of the conditions in Proposition \ref{prop:surjectivity}, and let 
\[
\alpha = \min\left(\prod_{i=1}^{n-1} \left\vert \sigma_i(M) \right\vert, 1\right).
\]
Then, for all $\rho \in (0, 1]$ and all
$r \in (0, \alpha^{-1} \cdot \rho^{-1})$, any lattice $H$ such that $B_V^-(\rho r) \subset H \subset B_V(r)$
satisfies:
\begin{equation}
\chi(M + H) = \chi(M) + d\chi_M(H).
\end{equation}
\end{prop}

@@

Recall \cite{caruso-roe-vaccon:15a}*{Def. 3.3} that the \emph{precision polygon}
of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X-M)$.
By \cite{caruso-roe-vaccon:15a}*{Prop. 3.4}, the endpoints of the precision polygon
occur at height $0$ and $\sum_{i=1}^{n-1} \val(\sigma_i(M))$.  By convexity,
$B_W(1) \subset d\chi(M)(B_V(\alpha^{-1}))$.

@@

\begin{proof}
Write $C = \com(X{-}M)$. From $(X{-}M) \cdot C \equiv 0 
\pmod{\chi_M}$, we deduce $(X{-}\mathscr{C}) \cdot P^{-1} C \equiv 0 \pmod{\chi_M}$. 
Therefore each column of $P^{-1} C$ lies in the right kernel of $X{-}\mathscr{C}$
modulo $\chi_M$. On the other hand, a direct computation shows that
every column vector $W$ lying in the right kernel of $X{-}\mathscr{C}$ modulo 
$\chi_M$ can be written as $W = w \cdot V^t$ for some $w \in 
K[X]/\chi_M$. We deduce that $C \equiv P \cdot V^t B \pmod{\chi_M}$
for some row vector $B$.
Applying the same reasoning with $M^t$, we find that $B$ can be
written $B = \alpha V Q^t$ for some $\alpha \in K[X]/\chi_M$ and
we are done.
\end{proof}

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ entries to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Section \ref{sec:diffFrob} that interesting
information can be read off of this short form $(\alpha, P, Q, 
\chi_M)$.

@@

In this section, we combine the computation of a Hessenberg form
of a matrix and the computation of the inverse through the Smith normal form (SNF)
over a complete discrete valuation field (CDVF)
to compute $\com(X-M)$ and $d \chi$.
If $M \in M_n(O_K)$, then only division by invertible
elements of $O_K$ will occur.
%The computation of a Hessenberg form provides an $O(n^3)$
%algorithm to compute the characteristic polynomial.

@@

We begin with the computation of
an approximate Hessenberg form.

\begin{deftn}
A \emph{Hessenberg matrix} is a matrix $M \in M_n(K)$ with
\[
M_{i,j}=0 \mbox{ for $j \le i-2$.}
\]
Given integers $n_{i,j}$, an \emph{approximate Hessenberg matrix}
is a matrix $M \in M_n(K)$ with
\[
M_{i,j} = O(\pi^{n_{i,j}}) \mbox{ for $j \le i-2$.}
\]
If $M \in M_n (K)$ and $H \in M_n (K)$ is an (approximate) Hessenberg matrix
similar to $M$, we say that H is an \emph{(approximate) Hessenberg form} of $M.$
\end{deftn}

It is not hard to prove that every matrix over a field admits
a Hessenberg form.
We prove here that over $K,$ if a matrix is 
known at finite (jagged) precision,
we can compute an approximate Hessenberg form of it.
Moreover, we can provide an exact change of basis matrix.
It relies on the following algorithm.

@@

\noindent 2.\  \:  {\bf swap} the row $j+1$ with a row $i_{min}$ ($i_{min} \geq 2$) s.t. $\val(H_{i_{min},j})$ is minimal.

@@

In this section, we prove that to compute the inverse of
a matrix over a CDVF $K$, SNF is precision-wise optimal in the flat-precision case.
We first recall what is the differential of the computation of the inverse of a matrix.

@@

In this subsection, we want to compute $\com (X Id -H)$
by using the SNF computation of the previous subsection.
As it is, in a way, optimal for the computation of the
inverse in a CDVF, we hope to get reasonable behavior
for the computation of $\com (X-M)$ even though the coefficients
are not, \textit{a priori}, set to live in a CDVF.
We show that we can reduce this task to the computation
of an SNF over a CDVF, with no division in $K.$

@@

\begin{rem}
The keyword \texttt{algorithm="df"} forces \texttt{SageMath} to use the 
division free algorithm of \cite{seifullin:02a}. It is likely that, proceeding so, we 
limit the loss of precision.
\end{rem}
